---
title: "Coronavirus Data Analysis"
author: "Marcelo Sanches"
date: "07/03/2022"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


This is an old project from the beginning of the pandemic which I'm revising by creating this [Shiny app.](https://bigbangdata.shinyapps.io/shinyapp/)

---

__April 2020__

This is a simple exploration of the time series data compiled from various sources at the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19). 

The raw data can also be downloaded manually at [Novel Coronavirus 2019 Cases.](https://data.humdata.org/dataset/novel-coronavirus-2019-ncov-cases) This [GitHub repository](https://github.com/BigBangData/CoronavirusDataAnalysis) hosts all files and code.


This project is not intended to be a serious data analysis, which would require more data and study. This is a personal project to explore automated plotting of the daily JHU datasets. **The plots produced here do not reflect reality and should not be taken as a model for how COVID-19 spreads through populations.** For example, they do not take into consideration the meaning of confirmed cases - this varies per location and time, availability of testing, changes in policy, and so forth. The data also might contain reporting errors.

---


## Contents {#contents-link}

* [Data Pre-Processing](#preprocess-link): brief description of data pre-processing and cleanup steps.
* [Data Wrangling and Enrichment](#enrich-link): adding population data and calculated columns.
* [Exploratory Data Analysis](#eda-link): main section with visualizations.
* [Code Appendix](#codeappendix-link): entire R code.

---

## Data Pre-Processing {#preprocess-link}

I focused on confirmed cases and fatal cases. See [Code Appendix](#codeappendix-link) for full, commented code.


```{r include=FALSE}
# setup
source("./00_setup_and_load.R")
rds_file <- paste0("./data/", gsub("-", "", Sys.Date()), "_data.rds")
dfm <- readRDS(rds_file)
```


```{r echo=FALSE}
# calculate number of countries and number of days in the time series
Ncountries <- length(unique(dfm$Country))
Ndays <- length(unique(dfm$Date))
```

### Summary

The pre-processed dataset is comprised of `r nrow(dfm)` rows and `r length(dfm)` columns. Each single-status dataset is as long as the number of days times the number of countries for the data in a given day. Today there are `r Ndays` days and `r Ncountries` countries in the data, after removing the small and seasonal populations of Antarctica and the Olympics.

The project focuses on countries so latitude, longitude, and the sub-national province/state columns were discarded. 


---

[Back to [Contents](#contents-link)]{style="float:right"}


## Data Wrangling and Enrichment {#enrich-link}


I maintain a static data set of countries and their populations. This data is cobbled together with internet searches and [World Health Organization data.](https://apps.who.int/gho/data/view.main.POP2040ALL?lang=en) I use the country's population to calculate two columns: `Count Per 10K` and `New Cases Per 10K` which track the cumulative cases were observed per 10,000 people, and how many new cases.

The top rows of the enriched data set for Brazil and the US are:

```{r echo=FALSE}
# top and bottom rows for final data set
kable(rbind(head(merged[merged$Country == "Brazil", ], 3)
            , head(merged[merged$Country == "US", ], 3))) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")
                  , full_width = FALSE)
```

---

[Back to [Contents](#contents-link)]{style="float:right"}


## Exploratory Data Analysis {#eda-link}


### Total Counts

```{r echo=FALSE, fig.height=6, fig.width=6} 
# subset to current counts 
current_data <- data.frame(
    merged %>%
    filter(Date == unique(merged$Date)[1])) %>%
    arrange(Status, desc(Count)
)

# subset to world totals 
world_totals <- data.frame(current_data %>% 
					group_by(Status) %>%
					summarise('Total'=sum(Count)))

world_totals$Total <- formatC(world_totals$Total, big.mark=",")

kable(world_totals) %>%
      kable_styling(bootstrap_options = c("striped", "hover")
                    , full_width = FALSE)
```



---

### Bar Charts

See this [Shiny app](https://bigbangdata.shinyapps.io/shinyapp/) for interactive bar charts with the latest counts and percentages.


```{r echo=FALSE}
# subset to country totals
country_totals <- data.frame(
    current_data %>%
		select(Country, Status, Count, Count_per10K, NewCases_per10K) %>%
		group_by(Country, Status)
)
	
# subset to top counts 	
get_top_counts <- function(dfm, coln, num) {
	  dfm <- dfm[dfm$Status == coln, ][1:num, ]
	  row.names(dfm) <- 1:num
	  dfm
}					

# separate by status 
top_confirmed 	<- get_top_counts(country_totals, "Confirmed", 10)
top_fatal		<- get_top_counts(country_totals, "Fatal", 10)
```


---

### Interactive Time Series Plots

Counts and percentages of the top five countries for confirmed and fatal cases.


```{r message=FALSE, warnings=FALSE, echo=FALSE}
# functions for plotting interactive time series
# arg values:
# dfm = the dataframe
# country = country name
# status_df = to be used as the vector of country names 
#             which is passed instead of a single country
# status = Confirmed, Fatal, Recovered, Active
# scale_ = Linear, Log
# type = Count, Percentage, NewCases
create_xts_series <- function(dfm, country, status, scale_, type) {
  
  	dfm <- dfm[dfm$Country == country & dfm$Status == status, ]
  	
  	if (type == "Count") {
  	  series <- if (scale_ == "Linear") {
  	  			  xts(dfm$Count, order.by = dfm$Date)
  	  		} else {
  	  		    xts(log(dfm$Count), order.by = dfm$Date)
  	  		}
  	} else if (type == "Percentage") {
  	  series <- if (scale_ == "Linear") {
  	  			  xts(dfm$Percentage, order.by = dfm$Date)
  	  		} else {
  	  		    xts(log(dfm$Percentage), order.by = dfm$Date)
  	  		}
  	} else { # for new cases
  	  series <- if (scale_ == "Linear") {
  	  			  xts(dfm$NewCases, order.by = dfm$Date)
  	  		} else {
  	  		    xts(log(dfm$NewCases), order.by = dfm$Date)
  	  		}
  	}
  	series
}

create_seriesObject <- function(dfm, status_df, status, scale_, type) {
  
    seriesObject <- NULL
    
    for (i in 1:5) {
        seriesObject <- cbind(seriesObject
                              , create_xts_series(dfm
                                                  , status_df$Country[i]
                                                  , status
                                                  , scale_
                                                  , type))
    }
    
    names(seriesObject) <- status_df$Country[1:5]
    seriesObject
}

plot_interactive_df <- function(dfm, status_df, status, scale_, type) {
  
    seriesObject <- create_seriesObject(dfm
                  									  , status_df
                  									  , status
                  									  , scale_
                  									  , type)
    
    if (type == "Count") {
  
        txt_ <- if (scale_ == "Linear") {
    	  				"Count Of "
    	  			} else {
    	  			  "Log Count Of "
    	  			}			
  				
    } else if (type == "Percentage") {
        
        txt_ <- if (scale_ == "Linear") {
    	  				"Percentage Of "
    	  			} else {
    	  			  "Log Percentage Of "
    	  			} 		
  				
    } else { # for new cases
      
        txt_ <- if (scale_ == "Linear") {
    	  				"New "
    	  			} else {
    	  			  "Log Of New "
    	  			}  	
    }
    
    ylab_lab   <- paste0(txt_, status, " Cases")
    main_title <- paste0("Top Countries - ", txt_, status, " Cases")
    
    interactive_df <- dygraph(seriesObject, main = main_title) %>% 
  					dyAxis("x", drawGrid = FALSE) %>%							
  					dyAxis("y", label = ylab_lab) %>%
  					dyOptions(colors=brewer.pal(5, "Dark2")
  							, axisLineWidth = 1.5
  							, axisLineColor = "navy"
  							, gridLineColor = "lightblue") %>%			
  					dyRangeSelector() %>%
  					dyLegend(width = 750)
    
    interactive_df
}
```


```{r message=FALSE, warnings=FALSE, echo=FALSE}
## plot time series
#plot_types <- data.frame(
#    'Num' = 1:12
#    , 'Status' = c(rep("Confirmed", 6), rep("Fatal", 6))
#    , 'Type' = rep(c("Count", "Percentage", "NewCases"), each=2)
#    , 'Scale' = rep(c("Linear", "Log"), 6)
#)
#
## fatal
#fatal_plots <- lapply(1:3, function(i) plot_interactive_df(merged
#  							                     , top_fatal[1:5, ]
#  							                     , top_fatal$Status[i]
#  							                     , plot_types$Scale[i]
#  							                     , plot_types$Type[i]))
#		
#htmltools::tagList(fatal_plots)
#
## confirmed 
#confirmed_plots <- lapply(1:3, function(i) plot_interactive_df(merged
#      							                     , top_confirmed[1:5, ]
#      							                     , top_confirmed$Status[i]
#      							                     , plot_types$Scale[i]
#      							                     , plot_types$Type[i]))
#		
#htmltools::tagList(confirmed_plots)
```



### Further Plots to Consider


Future iterations could include other interesting metrics such as:

- Doubling rates
- Proportion of New Cases to Total Cases
- Percentage increase plus a horizontal line showing proportion of population to world population


---


[Back to [Contents](#contents-link)]{style="float:right"}

### Code Appendix {#codeappendix-link}

Fork or clone this [GitHub repository](https://github.com/BigBangData/CoronavirusDataAnalysis) with all files and code, including the code for the Shiny app.


```{r}
# uncomment to run, creates Rcode file with R code, set documentation = 1 to avoid text commentary
#library(knitr)
#options(knitr.purl.inline = TRUE)
#purl("CoronavirusDataAnalysis.Rmd", output = "Rcode.R", documentation = 1)
``` 