---
title: "Coronavirus Data Analysis"
author: "Marcelo Sanches"
date: "3/29/2020"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
# setup
rm(list = ls())
options(scipen=999)

install_packages <- function(package){
  
  newpackage <- package[!(package %in% installed.packages()[, "Package"])]
      
	if (length(newpackage)) {
      suppressMessages(install.packages(newpackage, dependencies = TRUE))
	}
	sapply(package, require, character.only = TRUE)
}


# install packages  
packages <- c("dygraphs", "tidyverse", "xts", "RColorBrewer","kableExtra")
suppressPackageStartupMessages(install_packages(packages))
```


```{r include=FALSE}

# preprocessing function
preprocess <- function() {

	# create a folder for the data 
	dir_name <- "COVID19_DATA"
	if (!file.exists(dir_name)) {
		dir.create(dir_name)
	}
	
	dir_path <- "COVID19_DATA/"
	
	# download today's file, save as RDS first time, read otherwise
	file_name <- paste0(dir_path, gsub("-", "", Sys.Date()), "_data.rds")
	
	if (!file.exists(file_name)) {

		# create URLs
		http_header <- "https://data.humdata.org/hxlproxy/data/download/time_series_covid19_"
		
		url_body <- paste0("_narrow.csv?dest=data_edit&filter01=explode&explode-header-att01="
		                  ,"date&explode-value-att01=value&filter02=rename&rename-oldtag02=%23"
		                  ,"affected%2Bdate&rename-newtag02=%23date&rename-header02=Date&filter"
		                  ,"03=rename&rename-oldtag03=%23affected%2Bvalue&rename-newtag03=%23af"
		                  ,"fected%2Binfected%2Bvalue%2Bnum&rename-header03=Value&filter04=clea"
		                  ,"n&clean-date-tags04=%23date&filter05=sort&sort-tags05=%23date&sort-"
		                  ,"reverse05=on&filter06=sort&sort-tags06=%23country%2Bname%2C%23adm1%"
		                  ,"2Bname&tagger-match-all=on&tagger-default-tag=%23affected%2Blabel&t"
		                  ,"agger-01-header=province%2Fstate&tagger-01-tag=%23adm1%2Bname&tagger"
		                  ,"-02-header=country%2Fregion&tagger-02-tag=%23country%2Bname&tagger-"
		                  ,"03-header=lat&tagger-03-tag=%23geo%2Blat&tagger-04-header=long&tagge"
		                  ,"r-04-tag=%23geo%2Blon&header-row=1&url=https%3A%2F%2Fraw.githubuserc"
		                  ,"ontent.com%2FCSSEGISandData%2FCOVID-19%2Fmaster%2Fcsse_covid_19_data"
		                  ,"%2Fcsse_covid_19_time_series%2Ftime_series_covid19_")
		
		
		confirmed_URL  <- paste0(http_header, "confirmed_global", url_body, "confirmed_global.csv")
		fatal_URL <- paste0(http_header, "deaths_global", url_body, "deaths_global.csv")
		recovered_URL  <- paste0(http_header, "recovered_global", url_body, "recovered_global.csv")
									
		# download
		download.file(confirmed_URL, destfile=paste0(dir_path, "confirmed.csv"))
		download.file(fatal_URL, destfile=paste0(dir_path, "fatal.csv"))
		download.file(recovered_URL, destfile=paste0(dir_path, "recovered.csv"))
		
		# load csvs
		load_csv <- function(filename) { 
			filename <- read.csv(paste0(dir_path, filename, ".csv"), header=TRUE
			                     , fileEncoding="UTF-8-BOM"
								           , stringsAsFactors=FALSE, na.strings="")[-1, ]
			filename
		}
	
		confirmed  <- load_csv("confirmed")
		fatal <- load_csv("fatal") 
		recovered  <- load_csv("recovered")
		
		# prep data for long format
		
		# add column identifying the dataset	
		add_col <- function(dfm, name) {
			dfm$Status <- rep(name, nrow(dfm))
			dfm
		}
		
		confirmed  <- add_col(confirmed, "confirmed")
		fatal <- add_col(fatal, "fatal")
		recovered  <- add_col(recovered, "recovered")
		
		# join (union actually) into one dataset 
		dfm <- rbind(confirmed, fatal, recovered, make.row.names=FALSE)
		
		# rename columns 
		colnames(dfm) <- c("Province_State", "Country_Region"
				  , "Lat", "Long", "Date", "Value", "Status")
		
		# fix data types 
		dfm$Value <- as.integer(dfm$Value)
		dfm$Lat <- as.numeric(dfm$Lat)
		dfm$Long <- as.numeric(dfm$Long)
		dfm$Date <- as.Date(dfm$Date)
		dfm$Status <- as.factor(dfm$Status)
	
		# save as RDS 
		saveRDS(dfm, file = file_name)
		
	} 

	dfm <- readRDS(file_name) 

}

```



This is a simple exploration of the time series data which was compiled by the Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE) from various sources (see website for full description). The data can be downloaded manually at [Novel Coronavirus 2019 Cases.](https://data.humdata.org/dataset/novel-coronavirus-2019-ncov-cases)


## Contents {#contents-link}

* [Data Pre-Processing](#preprocess-link)
* [Data Cleanup](#cleanup-link)
* [Exploratory Data Analysis](#eda-link)
* [Code Appendix](#codeappendix-link)

---

## Data Pre-Processing {#preprocess-link}

The `preprocess` function creates a local folder and pulls three csv files, one for each stage in tracking the coronavirus spread (confirmed, fatal, and recovered cases), performs various pre-processing steps to create one narrow and long dataset, saving it in compressed RDS format. See code in the [Code Appendix.](#codeappendix-link)



```{r}
# read in RDS file 
dfm <- preprocess()

str(dfm)
```


There are `r nrow(dfm)` rows and `r length(dfm)` columns. There's a 'Status' column for the different stages, so the number of rows is 3 times the number of rows for a single status (ex. "confirmed"). Each single-status dataset is as long as the number of days in the time series (for a given day the data is pulled) times the number of countries and sub-national provinces or states. This number varies per country, and also varies per day depending on how the dataset is built. 


---

[Back to [Contents](#contents-link)]{style="float:right"}

## Data Cleanup  {#cleanup-link}


### Location Granularity 

The data's location variables have several issues. I will discard `Lat` and `Long` since I'm not doing any mapping. The variables `Country_Region` and `Province_State` are often loosely aggregated. This can be visualized in [Johns Hopkins' dashboard](https://coronavirus.jhu.edu/map.html): the totals for fatalities are grouped by a mixture of countries and subnational geographic areas. The US is conspicuously missing as a country. 

Since subnational data is sparse, I'll focus on country-level data. After some data analysis, I noticed that the anomalies will repond to one simple aggregation and I recreated the dataset at this national level. Canada is a prime example of bad data: notice how it lacks subnational data on recovered cases, but also, I doubt there's a province in Canada called 'Recovered':


```{r echo=FALSE}
# Canada provinces example
kable(data.frame(dfm[dfm$Country_Region == "Canada", ]) %>% 
		   distinct(Country_Region, Province_State, Status)) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")
                  , full_width = FALSE)
```

```{r include=FALSE}
# country-level dataset
country_level_df <- data.frame(dfm %>%
							   select(Country_Region, Status, Date, Value) %>%
							   group_by(Country_Region, Status, Date) %>%
							   summarise('Value'=sum(Value))) %>%
							   arrange(Country_Region, Status, desc(Date))

colnames(country_level_df) <- c("Country", "Status", "Date", "Count")

Ncountries <- length(unique(country_level_df$Country))
Ndays <- length(unique(country_level_df$Date))

# check: is the number of rows equal to the number of countries
# times the number of days times 3 (statuses)?
nrow(country_level_df) == Ncountries * Ndays * 3
```

The top and bottom rows for the final dataset:

```{r echo=FALSE}
# top and bottom rows for final dataset
kable(rbind(head(country_level_df)
     ,tail(country_level_df))) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")
                  , full_width = FALSE)
```





---

[Back to [Contents](#contents-link)]{style="float:right"}


## Exploratory Data Analysis {#eda-link}

#### WORLD TOTALS



```{r echo=FALSE} 
# subset to current counts 
current_data <- data.frame(country_level_df %>%
					filter(Date == unique(country_level_df$Date)[1])) %>%
					arrange(Status, desc(Count))

# subset to world totals 
world_totals <- data.frame(current_data %>% 
					group_by(Status) %>%
					summarise('total'=sum(Count)))

kable(world_totals) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


#### TOP COUNTRIES PER STATUS

```{r echo=FALSE}
# subset to country totals 
country_totals <- data.frame(current_data %>%
						select(Country, Status, Count) %>%
						group_by(Country, Status))
	
# subset to top counts 	
get_top_counts <- function(dfm, coln) {
	
	dfm <- dfm[dfm$Status == coln, c(1,3)][1:6,]
	row.names(dfm) <- 1:6
	dfm
}					

# separate by status 
top_confirmed 	<- get_top_counts(country_totals, "confirmed")
top_fatal	<- get_top_counts(country_totals, "fatal")
top_recovered 	<- get_top_counts(country_totals, "recovered")

# plot top countries per status 
gg_plot <- function(dfm, status, color) {

	ggplot(data=dfm, aes(x=reorder(Country, -Count), y=Count)) +
		geom_bar(stat="identity", fill=color) + 
		ggtitle(paste0("Top ", status, " Cases by Country")) + 
		xlab("") + ylab(paste0("Number of ", status, " Cases")) +
		geom_text(aes(label=Count), vjust=1.6, color="white", size=3.5) +
		theme_minimal()

}

# top confirmed
gg_plot(top_confirmed, "Confirmed", "red3")

# top fatal 
gg_plot(top_fatal, "Fatal", "gray25")

# top recovered
gg_plot(top_recovered, "Recovered", "springgreen4")

```




---

  
  

### Time Series Plots per Status and Location

This interactive time series speaks for itself: the US has overtaken Italy and China in number of confirmed cases in the last two days.



```{r fig.height=5, fig.width=9, echo=FALSE}
# function to create an xts series given dataframe, country, and status
create_xts_series <- function(dfm, country, status) {
  
	dfm <- dfm[dfm$Country == country & dfm$Status == status, ]
	series <- xts(dfm$Count, order.by = dfm$Date)
	series
}

# Confirmed
US <- create_xts_series(country_level_df, "US", "confirmed")
Italy <- create_xts_series(country_level_df, "Italy", "confirmed")
China <- create_xts_series(country_level_df, "China", "confirmed")
Spain <- create_xts_series(country_level_df, "Spain", "confirmed")
Germany <- create_xts_series(country_level_df, "Germany", "confirmed")

seriesObject <- cbind(US, Italy, China, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="US Overtakes Italy and China in Confirmed Cases"
						   ,xlab=""
						   ,ylab="Number of Confirmed Cases") %>% 
						   dyOptions(colors = brewer.pal(5,"Dark2")) %>%						  
						   dyRangeSelector()


dfm_interactive
```


This is the same visualization using the data on fatalities:

```{r fig.height=5, fig.width=9, echo=FALSE}
# Fatalities
US <- create_xts_series(country_level_df, "US", "fatal")
Italy <- create_xts_series(country_level_df, "Italy", "fatal")
China <- create_xts_series(country_level_df, "China", "fatal")
Spain <- create_xts_series(country_level_df, "Spain", "fatal")
Germany <- create_xts_series(country_level_df, "Germany", "fatal")

seriesObject <- cbind(US, Italy, China, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="Italy Leads in Fatalities"
						   ,xlab=""
						   ,ylab="Number of Fatalities") %>% 
						   dyOptions(colors = brewer.pal(5,"Dark2")) %>%						  
						   dyRangeSelector()

dfm_interactive
```


This is the same visualization using the data on recoveries:

```{r fig.height=5, fig.width=9, echo=FALSE}
# Recovered
US <- create_xts_series(country_level_df, "US", "recovered")
Italy <- create_xts_series(country_level_df, "Italy", "recovered")
China <- create_xts_series(country_level_df, "China", "recovered")
Spain <- create_xts_series(country_level_df, "Spain", "recovered")
Germany <- create_xts_series(country_level_df, "Germany", "recovered")

seriesObject <- cbind(US, Italy, China, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="China Leads in Recoveries"
						   ,xlab=""
						   ,ylab="Number of Recoveries") %>% 
						   dyOptions(colors = brewer.pal(5,"Dark2")) %>%						  
						   dyRangeSelector()

dfm_interactive
```

Since China dominates this plot too much, it would be interesting to see how the other countries are doing as far as recoveries:

```{r fig.height=5, fig.width=9, echo=FALSE}
# Recovered - other four countries
seriesObject <- cbind(US, Italy, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="After China, Italy and Spain Lead in Recoveries"
						   ,xlab=""
						   ,ylab="Number of Recoveries") %>% 
						   dyOptions(colors = brewer.pal(4,"Dark2")) %>%						  
						   dyRangeSelector()

dfm_interactive
```

#### Per Capita Analysis

```
ADD WORDS HERE

```

```{r include=FALSE}
# read in prepared dataset of countries and populations
country_population <- read.csv("COVID19_DATA/country_population.csv")
		  
# per capita analysis
percap <- merge(country_level_df, country_population, by="Country")

# percentage
percap$Pct <- (percap$Count / (percap$Population_thousands*1000)) * 100 
```


Top 25 Confirmed Cases by Percentage of Population

```{r echo=FALSE}
# subset to current counts 
current_data <- data.frame(percap %>%
					filter(Date == unique(percap$Date)[1])) %>%
					arrange(Status, desc(Pct))

kable(current_data[1:25, ]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```




```{r}	
# subset to top counts 	
get_top_pcts <- function(dfm, coln) {
	
	dfm <- dfm[dfm$Status == coln, c(1,6)][2:10,]
	row.names(dfm) <- 1:9
	dfm$Pct <- round(dfm$Pct, 4)
	dfm
}					

# separate by status 
top_confirmed 	<- get_top_pcts(current_data, "confirmed")
top_fatal	<- get_top_pcts(current_data, "fatal")
top_recovered 	<- get_top_pcts(current_data, "recovered")



# plot top countries per status 
gg_plot <- function(dfm, status, color) {

	ggplot(data=dfm, aes(x=reorder(Country, Pct), y=Pct)) +
		geom_bar(stat="identity", fill=color) + 
		ggtitle(paste0("Top ", status, " Cases by Pct of Population")) + 
		xlab("") + ylab(paste0("Percentage of ", status, " Cases")) +
		geom_text(aes(label=Pct), hjust=1.6, color="white", size=3.5) +
    coord_flip() + theme_minimal()

}

# top confirmed
gg_plot(top_confirmed, "Confirmed", "red3")

# top fatal 
gg_plot(top_fatal, "Fatal", "gray25")

# top recovered
gg_plot(top_recovered, "Recovered", "springgreen4")

```




---

[Back to [Contents](#contents-link)]{style="float:right"}

### Code Appendix {#codeappendix-link}

```{r eval=FALSE}
## ----setup, include=FALSE------------------------------------------------
knitr::opts_chunk$set(echo = TRUE)

## ----include=FALSE-------------------------------------------------------
# setup
rm(list = ls())
options(scipen=999)

install_packages <- function(package){
  
  newpackage <- package[!(package %in% installed.packages()[, "Package"])]
      
	if (length(newpackage)) {
      suppressMessages(install.packages(newpackage, dependencies = TRUE))
	}
	sapply(package, require, character.only = TRUE)
}


# install packages  
packages <- c("dygraphs", "tidyverse", "xts", "RColorBrewer","kableExtra")
suppressPackageStartupMessages(install_packages(packages))

## ----include=FALSE-------------------------------------------------------

# preprocessing function
preprocess <- function() {

	# create a folder for the data 
	dir_name <- "COVID19_DATA"
	if (!file.exists(dir_name)) {
		dir.create(dir_name)
	}
	
	dir_path <- "COVID19_DATA/"
	
	# download today's file, save as RDS first time, read otherwise
	file_name <- paste0(dir_path, gsub("-", "", Sys.Date()), "_data.rds")
	
	if (!file.exists(file_name)) {

		# create URLs
		http_header <- "https://data.humdata.org/hxlproxy/data/download/time_series_covid19_"
		
		url_body <- paste0("_narrow.csv?dest=data_edit&filter01=explode&explode-header-att01="
		                  ,"date&explode-value-att01=value&filter02=rename&rename-oldtag02=%23"
		                  ,"affected%2Bdate&rename-newtag02=%23date&rename-header02=Date&filter"
		                  ,"03=rename&rename-oldtag03=%23affected%2Bvalue&rename-newtag03=%23af"
		                  ,"fected%2Binfected%2Bvalue%2Bnum&rename-header03=Value&filter04=clea"
		                  ,"n&clean-date-tags04=%23date&filter05=sort&sort-tags05=%23date&sort-"
		                  ,"reverse05=on&filter06=sort&sort-tags06=%23country%2Bname%2C%23adm1%"
		                  ,"2Bname&tagger-match-all=on&tagger-default-tag=%23affected%2Blabel&t"
		                  ,"agger-01-header=province%2Fstate&tagger-01-tag=%23adm1%2Bname&tagger"
		                  ,"-02-header=country%2Fregion&tagger-02-tag=%23country%2Bname&tagger-"
		                  ,"03-header=lat&tagger-03-tag=%23geo%2Blat&tagger-04-header=long&tagge"
		                  ,"r-04-tag=%23geo%2Blon&header-row=1&url=https%3A%2F%2Fraw.githubuserc"
		                  ,"ontent.com%2FCSSEGISandData%2FCOVID-19%2Fmaster%2Fcsse_covid_19_data"
		                  ,"%2Fcsse_covid_19_time_series%2Ftime_series_covid19_")
		
		
		confirmed_URL  <- paste0(http_header, "confirmed_global", url_body, "confirmed_global.csv")
		fatal_URL <- paste0(http_header, "deaths_global", url_body, "deaths_global.csv")
		recovered_URL  <- paste0(http_header, "recovered_global", url_body, "recovered_global.csv")
									
		# download
		download.file(confirmed_URL, destfile=paste0(dir_path, "confirmed.csv"))
		download.file(fatal_URL, destfile=paste0(dir_path, "fatal.csv"))
		download.file(recovered_URL, destfile=paste0(dir_path, "recovered.csv"))
		
		# load csvs
		load_csv <- function(filename) { 
			filename <- read.csv(paste0(dir_path, filename, ".csv"), header=TRUE
			                     , fileEncoding="UTF-8-BOM"
								           , stringsAsFactors=FALSE, na.strings="")[-1, ]
			filename
		}
	
		confirmed  <- load_csv("confirmed")
		fatal <- load_csv("fatal") 
		recovered  <- load_csv("recovered")
		
		# prep data for long format
		
		# add column identifying the dataset	
		add_col <- function(dfm, name) {
			dfm$Status <- rep(name, nrow(dfm))
			dfm
		}
		
		confirmed  <- add_col(confirmed, "confirmed")
		fatal <- add_col(fatal, "fatal")
		recovered  <- add_col(recovered, "recovered")
		
		# join (union actually) into one dataset 
		dfm <- rbind(confirmed, fatal, recovered, make.row.names=FALSE)
		
		# rename columns 
		colnames(dfm) <- c("Province_State", "Country_Region"
				  , "Lat", "Long", "Date", "Value", "Status")
		
		# fix data types 
		dfm$Value <- as.integer(dfm$Value)
		dfm$Lat <- as.numeric(dfm$Lat)
		dfm$Long <- as.numeric(dfm$Long)
		dfm$Date <- as.Date(dfm$Date)
		dfm$Status <- as.factor(dfm$Status)
	
		# save as RDS 
		saveRDS(dfm, file = file_name)
		
	} 

	dfm <- readRDS(file_name) 

}


## ------------------------------------------------------------------------
# read in RDS file 
dfm <- preprocess()

str(dfm)


nrow(dfm)
length(dfm)
## ----echo=FALSE----------------------------------------------------------
# Canada provinces example
kable(data.frame(dfm[dfm$Country_Region == "Canada", ]) %>% 
		   distinct(Country_Region, Province_State, Status)) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")
                  , full_width = FALSE)

## ----include=FALSE-------------------------------------------------------
# country-level dataset
country_level_df <- data.frame(dfm %>%
							   select(Country_Region, Status, Date, Value) %>%
							   group_by(Country_Region, Status, Date) %>%
							   summarise('Value'=sum(Value))) %>%
							   arrange(Country_Region, Status, desc(Date))

colnames(country_level_df) <- c("Country", "Status", "Date", "Count")

Ncountries <- length(unique(country_level_df$Country))
Ndays <- length(unique(country_level_df$Date))

# check: is the number of rows equal to the number of countries
# times the number of days times 3 (statuses)?
nrow(country_level_df) == Ncountries * Ndays * 3

## ----echo=FALSE----------------------------------------------------------
# top and bottom rows for final dataset
kable(rbind(head(country_level_df)
     ,tail(country_level_df))) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")
                  , full_width = FALSE)

## ----echo=FALSE----------------------------------------------------------

# subset to current counts 
current_data <- data.frame(country_level_df %>%
					filter(Date == unique(country_level_df$Date)[1])) %>%
					arrange(Status, desc(Count))

# subset to world totals 
world_totals <- data.frame(current_data %>% 
					group_by(Status) %>%
					summarise('total'=sum(Count)))

kable(world_totals) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

## ----echo=FALSE----------------------------------------------------------


# subset to country totals 
country_totals <- data.frame(current_data %>%
						select(Country, Status, Count) %>%
						group_by(Country, Status))
	
# subset to top counts 	
get_top_counts <- function(dfm, coln) {
	
	dfm <- dfm[dfm$Status == coln, c(1,3)][1:6,]
	row.names(dfm) <- 1:6
	dfm
}					

# separate by status 
top_confirmed 	<- get_top_counts(country_totals, "confirmed")
top_fatal	<- get_top_counts(country_totals, "fatal")
top_recovered 	<- get_top_counts(country_totals, "recovered")

# plot top countries per status 
gg_plot <- function(dfm, status, color) {

	ggplot(data=dfm, aes(x=reorder(Country, -Count), y=Count)) +
		geom_bar(stat="identity", fill=color) + 
		ggtitle(paste0("Top ", status, " Cases by Country")) + 
		xlab("") + ylab(paste0("Number of ", status, " Cases")) +
		geom_text(aes(label=Count), vjust=1.6, color="white", size=3.5) +
		theme_minimal()

}

# top confirmed
gg_plot(top_confirmed, "Confirmed", "red3")

# top fatal 
gg_plot(top_fatal, "Fatal", "gray25")

# top recovered
gg_plot(top_recovered, "Recovered", "springgreen4")


## ----fig.height=5, fig.width=9, echo=FALSE-------------------------------
# function to create an xts series given dataframe, country, and status
create_xts_series <- function(dfm, country, status) {
  
	dfm <- dfm[dfm$Country == country & dfm$Status == status, ]
	series <- xts(dfm$Count, order.by = dfm$Date)
	series
}

# Confirmed
US <- create_xts_series(country_level_df, "US", "confirmed")
Italy <- create_xts_series(country_level_df, "Italy", "confirmed")
China <- create_xts_series(country_level_df, "China", "confirmed")
Spain <- create_xts_series(country_level_df, "Spain", "confirmed")
Germany <- create_xts_series(country_level_df, "Germany", "confirmed")

seriesObject <- cbind(US, Italy, China, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="US Overtakes Italy and China in Confirmed Cases"
						   ,xlab=""
						   ,ylab="Number of Confirmed Cases") %>% 
						   dyOptions(colors = brewer.pal(5,"Dark2")) %>%						  
						   dyRangeSelector()


dfm_interactive

## ----fig.height=5, fig.width=9, echo=FALSE-------------------------------
# Fatalities
US <- create_xts_series(country_level_df, "US", "fatal")
Italy <- create_xts_series(country_level_df, "Italy", "fatal")
China <- create_xts_series(country_level_df, "China", "fatal")
Spain <- create_xts_series(country_level_df, "Spain", "fatal")
Germany <- create_xts_series(country_level_df, "Germany", "fatal")

seriesObject <- cbind(US, Italy, China, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="Italy Leads in Fatalities"
						   ,xlab=""
						   ,ylab="Number of Fatalities") %>% 
						   dyOptions(colors = brewer.pal(5,"Dark2")) %>%						  
						   dyRangeSelector()

dfm_interactive

## ----fig.height=5, fig.width=9, echo=FALSE-------------------------------
# Recovered
US <- create_xts_series(country_level_df, "US", "recovered")
Italy <- create_xts_series(country_level_df, "Italy", "recovered")
China <- create_xts_series(country_level_df, "China", "recovered")
Spain <- create_xts_series(country_level_df, "Spain", "recovered")
Germany <- create_xts_series(country_level_df, "Germany", "recovered")

seriesObject <- cbind(US, Italy, China, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="China Leads in Recoveries"
						   ,xlab=""
						   ,ylab="Number of Recoveries") %>% 
						   dyOptions(colors = brewer.pal(5,"Dark2")) %>%						  
						   dyRangeSelector()

dfm_interactive

## ----fig.height=5, fig.width=9, echo=FALSE-------------------------------
# Recovered - other four countries
seriesObject <- cbind(US, Italy, Spain, Germany)
				 
dfm_interactive <- dygraph(seriesObject
						   ,main="After China, Italy and Spain Lead in Recoveries"
						   ,xlab=""
						   ,ylab="Number of Recoveries") %>% 
						   dyOptions(colors = brewer.pal(4,"Dark2")) %>%						  
						   dyRangeSelector()

dfm_interactive


```



```{r}
# uncomment to run, creates Rcode file with R code, set documentation = 1 to avoid text commentary
#library(knitr)
#options(knitr.purl.inline = TRUE)
#purl("COVID19_DATA_ANALYSIS.Rmd", output = "Rcode.R", documentation = 2)
``` 


